\documentclass{article}
\usepackage{iclr2026_conference}
\input{math_commands.tex}
\usepackage{booktabs}
\usepackage{url}

\title{Embodied Attention Budgeting}
\author{Anonymous Authors}
\date{}

\begin{document}
\maketitle

\paragraph{Submission-hardening version: v2.}
This revision adds a horizon stress. The budgeted policy wins at horizon 12
with 1,572/2,000 successes while greedy information gathering has 0/2,000, but
at horizon 16 greedy recovers to 1,886/2,000 successes versus the budgeted
policy's 1,572/2,000. The supported claim is therefore finite-horizon,
risk-coupled attention budgeting, not a universal rejection of information gain.

\begin{abstract}
Active perception usually asks which observation would be most informative.
For embodied robots, that question is incomplete: sensing consumes time,
physical configuration, compute, power, and safety margin. We propose
\emph{embodied attention budgeting}: a controller treats sensing as a limited
physical action and spends it only when the expected control-risk reduction
justifies the budget. A small hazard-navigation simulation shows why the
distinction matters. A no-sensing baseline reaches the goal often but collides
frequently. A greedy information policy over-senses and misses the finite
control horizon. A risk-coupled budgeted policy spends only two sensing actions,
reduces collisions, and keeps enough horizon to finish the task. The result is
not a broad benchmark claim; it is mechanistic evidence that attention should be
accounted for as an embodied resource, not treated as free uncertainty
reduction. A v2 horizon stress shows the boundary: when the horizon grows from
12 to 16 steps, greedy sensing recovers to 1,886/2,000 successes and exceeds the
budgeted policy.
\end{abstract}

\section{Introduction}
Robots do not observe from nowhere. A camera view can require motion, a tactile
measurement can require contact, and a high-resolution perception pass can cost
latency while the robot is still moving. Yet much active perception work frames
the problem as choosing the most informative observation. That framing hides a
physical assumption: sensing is cheap enough, safe enough, and separable enough
to optimize before or outside control.

This paper asks a narrower question. What if attention itself is a scarce
embodied resource? We use \emph{attention} broadly: visual fixation, tactile
probing, active view selection, high-resolution inference, or any sensing action
that consumes time and physical opportunity. The contribution is not a larger
model, a new benchmark, or reinforcement learning. It is a change in accounting.
The robot should spend attention only when the sensing action preserves more
control value than it consumes.

Our claims are modest. First, information gain alone can overvalue sensing when
there is a finite control horizon. Second, a risk-coupled budget can select
less information and still produce better task behavior. Third, the difference
is observable in a small runnable task with a fixed hazard, a finite horizon,
and a limited attention budget.

\section{Related Work}
The closest areas are active perception, active sensing, sensor scheduling,
information gathering, tactile exploration, and perception-aware manipulation.
The 2,013-row sweep in \texttt{docs/related\_work\_matrix.csv} covers robotics
and robotics-adjacent arXiv papers, then narrows to active-perception and
sensor-management clusters. The strongest hostile priors include active
sensing surveys, multi-robot information acquisition, attention-based planning
with active perception, submodular active perception, and active neural sensing.

These works make it non-novel to choose views, maximize information, schedule
sensors, or couple perception to action. The remaining gap is the objective.
Many methods ask whether an observation is informative. Fewer ask whether the
robot can afford the embodied act of attending under immediate control risk.

\section{Budgeted Attention}
Let $s_t$ be the robot state, $b_t$ its belief uncertainty, $u_t$ a control
action, and $o_t$ a sensing action. A standard active-perception objective can
choose $o_t$ by maximizing expected information gain. We instead introduce an
attention budget $B_t$ and a risk-coupled spending rule:
\[
  o_t \text{ is allowed only if }
  \Delta R(s_t,b_t,o_t) > C_{\mathrm{attn}}(o_t, B_t, H_t),
\]
where $\Delta R$ is expected control-risk reduction, $C_{\mathrm{attn}}$ is the
embodied cost of sensing, and $H_t$ is the remaining control horizon. The
important feature is not the exact formula; it is that attention competes with
actuation for finite physical opportunity.

\section{Toy Evidence}
We implement a one-dimensional hazard task in
\texttt{scripts/toy\_attention\_budget.py}. The robot must reach a target at
position 10 within 12 steps while crossing a hazard near position 5. Moving
under uncertainty can collide. Sensing reduces uncertainty, but it consumes one
step and one unit of attention.

The policies are deliberately simple. The baseline never senses. The greedy
policy senses until uncertainty is very low. The budgeted policy senses only
while uncertainty is high and the robot is not at the risky hazard boundary.
All policies are evaluated for 2,000 deterministic seeds.

\begin{table}[h]
\centering
\begin{tabular}{lrrrrr}
\toprule
Policy & Success & Collisions & Avg. return & Attention & Info \\
\midrule
Baseline & 836 & 1164 & -1.734 & 0.000 & 0.000 \\
Greedy & 0 & 114 & -1.767 & 5.000 & 3.564 \\
Budgeted & 1572 & 428 & 1.049 & 2.000 & -0.044 \\
\bottomrule
\end{tabular}
\caption{Toy attention-budget task. Greedy sensing gathers more information but
spends too much horizon to finish. Budgeted attention spends less and succeeds
more often.}
\end{table}

The exact values are less important than the failure mode. Information gain is
not the right objective when sensing consumes the same finite horizon needed
for safe control.

\paragraph{V2 horizon stress.}
The main table depends on a tight 12-step control horizon. To attack that
assumption, we rerun the same deterministic seeds with horizons 12, 14, 16, and
18. Table~\ref{tab:horizon} shows that greedy information gathering is bad only
when its five sensing actions consume too much of the horizon. At horizon 16,
greedy reaches 1,886 successes with 114 collisions and average return 1.807,
while the budgeted policy remains at 1,572 successes and average return 1.049.
Thus the claim is not that information gain is wrong. The claim is that
information gain must be priced by remaining control opportunity.

\begin{table}[h]
\centering
\input{docs/toy_attention_horizon_stress_table.tex}
\caption{V2 horizon stress. When the control horizon is long enough, greedy
information gathering recovers and outperforms the budgeted policy.}
\label{tab:horizon}
\end{table}

\section{Limitations}
This is a toy experiment, not a hardware result. The budget rule is hand-coded,
the hazard is simple, and the attention cost is stylized. The claim is also
close to prior work in active perception with costs. The novelty boundary is
therefore narrow: attention is treated as a physically consumed control
resource, not merely as an observation-selection cost. The v2 stress makes the
boundary explicit: if the horizon is long enough or sensing is cheap enough,
greedy information gathering can be the better policy.

\section{Conclusion}
Embodied attention should be budgeted. In robots, sensing can improve belief
while still damaging control opportunity. A small simulation shows that a
risk-coupled budget can outperform both no sensing and greedy sensing. The next
step is to validate the same accounting on real perception-action loops where
view changes, tactile probes, and compute latency have measurable physical
cost.

\section*{References}
\small
Active sensing for robotics surveys; decentralized active information
acquisition; attention-based planning with active perception; submodular active
perception; multi-modal active perception for information gathering; active
perception and representation for robotic manipulation; robot active neural
sensing and planning; tactile and visuo-tactile active perception; and
perception-aware robot navigation and manipulation papers identified in
\texttt{docs/hostile\_prior\_work.md}.

\end{document}
